### 2004-07-19 Documentation/part2.tex
Editing done ;-)

\part{The Abstract Experiment Definition Language -- AEDL}
\label{part:2}

@@

The goal of AEDL is to describe a large number of complex and very 
different experiments by a limited number of parameters. It allows a
representation of very different setups within one data structure, and thus implements universal rate and $\chi^2$ computation methods. For experiment simulations, a new piece of code is usually written and compiled
for each different experiment. In many cases, even parameter changes, such as
the number of bins, requires the recompilation of the source code. 
However, such a technique soon reaches its limits when the simulated experiments are rather complex, or more than one type of experiment is studied simultaneously. Furthermore, it is very difficult to verify the correctness of the obtained results, since every time a new piece of code is added to 
deal with a new experiment type, new errors will be introduced.

Thus, a general and flexible experiment description language is needed.  
The description of a neutrino experiment can be split into three parts: Source, oscillation, and detection. The neutrino sources within \GLOBES\ 
are assumed to be stationary point sources, where each experiment has only 
one source. This restricts the classes of neutrino sources which can be studied with \GLOBES :
\begin{itemize}
\item
 Experiments using many point-like sources can only be approximated. One example are reactor experiments using many distant reactor blocks.
\item
 Geometrical effects of a source distribution, such as in the sun or the atmosphere, can not be described.
\item
 Sources with a physically significant time dependencies  can not be studied, such as  supernov\ae. It is, however, possible
to study beams with bunch structure, since the time dependence of the
neutrino source is physically only important to suppress backgrounds. 
\end{itemize}

The description of the neutrino oscillation physics is, at least numerically, relatively simple. We use the {\em evolution operator method} to compute the neutrino oscillation probabilities and divide the matter density profile into layers with constant matter densities. For each of these layers, the Hamiltonian in matter is diagonalized in order to propagate the neutrino transition amplitudes. Finally, the transition probability is obtained by the absolute square of the total neutrino transition amplitudes. Depending on the precision of the studied experiment, this approach turns out to be precise enough in Earth matter even if only a small number matter density steps is used. Since we allow an uncertainty of the matter density profile, it is, in fact, in most cases sufficient to consider only one density step with the average matter density together with a matter density uncertainty~\cite{Ohlsson:2003ip}. Note that this approach may not be applicable to quickly varying extraterrestrial matter density profiles.

@@

Experiments are in \GLOBES\ defined by the Abstract Experiment Definition Language (AEDL). The experiment definition is written into a text file using the AEDL syntax. Currently, a number of pre-defined experiment definition files are provided with \GLOBES , which have to be modified manually in order to define new experiments. An IDE (Integrated Developement Environment) for AEDL files may follow in one of the next
versions. The application software then uses this text file to initialize the experiment, where other secondary files might read for source fluxes, cross sections \etc . In this section, we show the definition of a very simple neutrino factory in AEDL, where we do not go into details. In the next chapter, we will discuss each of the individual steps in detail.

@@

originating from stored $\mu^+$'s. Therefore, we set the {\tt @builtin} variable to $1$. Next, we specify the muon energy to be $50\,\mathrm{GeV}$ by the {\tt @parent\_energy} variable. We assume 

@@

{\tt \tb @errordim = 0}\\

@@

simultaneously. It is only useful if there is more than one background component, otherwise it is usually $1$. The background error variable is defined such as the signal error variable, \ie\ we have a $5\%$ background uncertainty and a very small energy calibration error. The ``error dimension variable'' {\tt @errordim} selects how the systematical errors are treated (\cf, \Tab~\ref{tab:error_dim}).

@@

We now give a short introduction to the syntax of AEDL.
 The first eight characters have to be {\tt \%!GLoBES}
in order to avoid parsing megabytes of chunk
 and producing thousands of error messages. 
%
Comments can be used such as in C:
\begin{quote}
{\tt /* This starts a comment\\
 and here the comment ends */
}
\end{quote}
There are pre-defined variables which all start with {\tt \$}. Their range
is also checked. For example,  {\tt 
\$bins} can be only between $0$ and $500$.\footnote{The upper limit is 
only there for safety reasons, the memory is allocated dynamically.} If one uses a {\tt float} quantity where  an {\tt int} is expected, the {\tt float} will be converted to an {\tt int} in the same way as in C.  For example, we have scalar variables
\begin{quote}
{\tt
\$bins = 10\\
\$baseline = 1200.0
}
\end{quote}
and simple lists
\begin{quote}
{\tt
\$densitytab=\{1.0,2.2343,3.3432\} 
}
\end{quote}
%
Since there are often groups of data which we want to refer to later,
environments can be used. This is illustrated 
with the channel definition part:
\begin{quote}
{\tt channel(\#ch1)<\\
\tb  $\ldots$\\
>
}
\end{quote}
The first part is the type of environment, which is {\tt channel} here. 
There are the following types of environment in AEDL:
\begin{quote}
{\tt flux\\
cross\\
channel\\
energy\\
rule
}
\end{quote}
Besides the environment type, there is a user-defined name 
beginning with {\tt \#}
in the above example: {\tt \#ch1}. It can be used later to refer to the 
channel defined in {\tt <$\ldots$>}. Those names are so-called 
``automatic variables'' and have to start with {\tt \#}. Note that these names have to be unique and can only be refered to after their definition.
However, similar to C, one can give a header definition before:
\begin{quote}
{\tt    channel(\#ch2)<>}
\end{quote}
Now one can refer to the name {\tt \#ch2}, while the actual channel definition comes later. The internal representation of this automatic
variable is a number, which obtains its value from a counter for each type of environment. For example, for {\tt channel} the counter is {\tt numofchannels}. The counter keeps track of how many different names 
there are for one type of environment, which means that it counts the number of channels, rules, energy resolution functions \etc . Thus, the automatic
variables are numbered in the order of their definition, and the number
can later be used to refer to them in the C code (from $0$ to {\tt numof...}$-1$). Within each environment type, there are several 
variables beginning with {\tt @}, which can only be used within the 
appropriate type of environment. In many cases, 
they have a special syntax, such as {\tt @channel}.

\index{aedl}{NEXT@{\tt \#NEXT\#}}
If you want to have several experiments in one file, separate the different
 experiments by 
\begin{quote}
{\tt    \#NEXT\#}
\end{quote}
This command resets the counters for channels, rules, fluxes, and energy resolution environments. All variables have their scope limited 
by either {\tt \%!GLoBES, \#NEXT\#} or {\tt EOF}.  This allows 
a consistent treatment of various experiments in one file.

\index{aedl}{inlcude@{\tt include}}
As another feature of AEDL one can use include files with the {\tt include} command. Includes can be nested up to {\tt MAX\_INCLUSION\_DEPTH}, which is currently set to $10$. Error reporting works 
 for nested includes, too. The included file is not required to begin 
 with {\tt \%!GLoBES} to facilitate cut and paste:
\begin{quote}
{\tt include "./file\_1"}
\end{quote}
With this include mechanism, one can use constructions such as 
\begin{quote}
{\tt    include  "NuFact.gls"\\
        \#NEXT\#\\
        include   "JHFHK.gls"
}
\end{quote}
in order to initialize a combined analysis of the experiments defined in the files {\tt NuFact.gls} and {\tt JHFHK.gls}. Note that one has 
to use quotation marks for filenames in \AEDL.\index{norm}{File names}
Even if one uses the
automatic variable {\tt \#CC} in both experiments, 
but the cross section data are different (for example, because of different target nuclei), the correct 
cross section data will be applied to each of the experiments. 
Note that, alternatively, one can 
also load both files successively.

Furthermore, one can define constants such as
\begin{quote}
{\tt
Pi = 3.141
}
\end{quote}
These constants can not only be defined within one \AEDL\ file, but also
by the calling C program, which allows to use a simple but powerful variable
substitution mechanism as described in~\Sec~\ref{sec:aedlparams}.
\index{norm}{External \AEDL\ parameters}

@@

As we have disussed before, \GLOBES\ can only deal with point sources. Thus,  it is not possible to study effects from the finite size of the neutrino production region, such as in the Sun or in reactor experiments with many
neutrino sources (\eg, KamLAND). Therefore, a neutrino source in \GLOBES\ can, in general, be characterized by the flux spectrum for each neutrino flavor, the CP sign (neutrinos or antineutrinos), and the total luminosity
of the source.

Before we come to the definition of the source properties, let us discuss
the total integrated luminosity of the experiment. In \GLOBES , the total number of events is in general proportional to the product of
\begin{equation}
\mathrm{Fid.~detector~mass}\,\left[\mathrm{kt/t}\right]\times 
\mathrm{Running~time} \,\left[\mathrm{yr}\right]\times\left\{ \begin{array}{c}
\mathrm{Source~power}\,\left[\mathrm{MW/GW}\right]\\
\mathrm{Useful~muon~decays}\,\left[\mathrm{yr}^{-1}\right]
\end{array}\right.\,.
\end{equation}
Thus, the source power corresponds to either the amount of energy produced per time frame in the target (such as for nuclear reactors or sources based on pion decay), or the useful muon decays per time frame (neutrino factories). In addition, the definition of the source power makes only sense together with the flux normalization, the running time, and fiducial detector mass in order to define the total integrated luminosity. Therefore, one can, in principle, use arbitrary units for these components as long as their product gives the wanted neutrino flux. However, it is
recommended to use normalizations such that the source power units are $\mathrm{MW}$ for a proton-based beam, and $\mathrm{GW}_\mathrm{thermal}$ for a reactor experiment. Correspondingly, the detector mass units should be kilotons for a proton-based beam, and tons for a reactor experiment.

@@

\index{aedl}{{\tt flux}!flux file@{\tt "@flux\_file}}
The software assumes that the given flux file has seven columns and
501 lines with equidistant energies. The format is:
\begin{quotation}
$ E\quad
\Phi_{\nu_e}\quad
\Phi_{\nu_\mu}\quad
\Phi_{\nu_\tau}\quad
\Phi_{\bar\nu_e}\quad
\Phi_{\bar\nu_\mu}\quad
\Phi_{\bar\nu_\tau}$
\end{quotation}
In order to access fluxes at arbitrary energies, linear interpolation 
is used by the software. In general, it is advisable to provide the flux between {\tt \$sampling\_min} and {\tt \$sampling\_max} (\cf, \Sec~\ref{sec:energy}), since these values are used by the software. However, if the energy leaves the range of values given in the file, zero is returned. 
Note that unused (and therefore often unknown) fluxes can be initialized with zeros, but can not just be omitted. For example, if one wants to define an experiment with different running times with neutrinos and antineutrinos, one can just use the same flux file in two different {\tt flux} environments with two different running times {\tt @time}. The unused fluxes in each case will then just be ignored.

% Continue here (WW)

@@

Furthermore, the matter density profile along the baseline
has to be specified. The simplest matter density profile is a constant matter density profile with the average matter density from the PREM\cite{Stacey} onion shell model of the Earth\index{norm}{PREM}\index{norm}{Earth matter density}:
\begin{quote}
{\tt \$profiletype=1 }
\end{quote}
%
For a better approximation of the realistic Earth matter density profile, one can use an arbitrary number of equidistant steps of the PREM profile:
\begin{quote}
{\tt \$profiletype=2 } \\
{\tt \$densitysteps=20 }
\end{quote}
Note that the value of {\tt \$densitysteps} is time-critical, since the
computation time of oscillation probabilities is directly 
proportional to the number of layers.
%
As a  third possibility, one can specify the matter density profile 
manually with a list of thicknesses and densities of the matter density layers. This example uses two density steps with two different densities:
\begin{quote}
{\tt \$profiletype=3 } \\
{\tt \$densitytab=\{2.8, 3.5\}}\\
{\tt \$lengthtab=\{1000.0, 2000.0\}}
\end{quote}
It is important that both lists habe the same length and that the  thicknesses given in  {\tt \$lengthtab} add up to the length of
the baseline, which does not have to be explicitely specified anymore. In addition, matter densities are always given in $g/cm^3$.
%
This approach can also be used for a constant matter density profile with
a specific matter density:
\begin{quote}
{\tt \$profiletype=3 } \\
{\tt \$densitytab=\{3.5\}}\\
{\tt \$lengthtab=\{3000.0\}}
\end{quote}
The possible options for matter density profiles are summarized in \Tab~\ref{tab:profiletypes}.

@@

Cross sections\index{norm}{Cross sections} will later be used as part of the 
channel definition (see \Sec~\ref{sec:channel}). Similar to the source 
fluxes, they are provided by the user as a data file:
\index{aedl}{{\tt cross}}
\index{aedl}{{\tt cross}!cross file@{\tt  "@cross\_file}}
\begin{quote}
{\tt cross(\#name)<}\\
{\tt \tb @cross\_file ="user\_file\_1.dat"}\\
{\tt >}
\end{quote}  
This cross section can later be refered to by {\tt \#name}.

Cross sections are in \GLOBES\ given as differential cross section per energy:
\begin{equation}
\hat\sigma(E)=\sigma(E)/E\,\left[ 10^{-38}\,
\frac{\mathrm{cm}^2}{\mathrm{GeV}^2} \right]
\end{equation}
The software assumes that the cross section files are text files with 
seven columns and $1001$ lines of the form
\index{aedl}{{\tt cross}!cross file@{\tt  "@cross\_file}}
\begin{quotation}
$\mathrm{log}_{10} E\quad
\hat\sigma_{\nu_e}\quad
\hat\sigma_{\nu_\mu}\quad
\hat\sigma_{\nu_\tau}\quad
\hat\sigma_{\bar\nu_e}\quad
\hat\sigma_{\bar\nu_\mu}\quad
\hat\sigma_{\bar\nu_\tau}$
\end{quotation}
Here the logarithms of the energy values have to be equidistant. For 
arbitrary energies, linear interpolation is used. If the energy leaves the
range of values given in the file, $0.0$ will be assumed. In general, it is advisable to provide the cross sections in the range between {\tt \$sampling\_min} and {\tt \$sampling\_max} (\cf, \Sec~\ref{sec:energy}).
Unused cross sections have to be filled with zeros, and can not
 be just omitted.

@@

\index{aedl}{{\tt channel}|(}
Channels\index{norm}{Channel} in \GLOBES\ represent an intermediate level 
between the pure oscillation physics given by the oscillation probability
$P_{\alpha\beta}$, and the total event rates composed of signal and 
background. A channel describes the path from one
initial neutrino flavor in the source to the event rates in the detector for one specific interaction type (IT) and final flavor.  
Therefore, a channel contains the description of the 
initial neutrino flavor, its CP eigenvalue (neutrino or antineutrino), 
the detected neutrino flavor, the interaction cross sections for the chosen interaction type, and the energy resolution function of the detector.

@@

Since it is a lot of effort to solve this double integral numerically,
we split up the two integrations. First, we evaluate the integral over
$\hat{E}$, where the only terms containing $\hat{E}$ are
$k_\beta^{\text{IT}}(E-\hat{E})$,  $ T_\beta(\hat{E})$, and 
$ V_\beta(\hat{E}-E')$. We define:
\begin{eqnarray}
\label{eq:e_res} 
R_\beta^{\text{IT}}(E,E')\,\epsilon_\beta^{\text{IT}}(E')
 \equiv
\int\limits_0^\infty d\hat{E} \quad T_\beta(\hat{E})\,k_\beta^{\text{IT}}(E-\hat{E})
\,V_\beta(\hat{E}-E')\,. 
\end{eqnarray}
Thus, $R_\beta^{\text{IT}}(E,E')$ describes the energy response of 
the detector, \ie , a neutrino with a (true) energy $E$ is reconstructed
with an energy between $E'$ and $E'+dE'$ with a probability
$R_\beta^{\text{IT}}(E,E') dE'$. The function $R(E,E')$ is also often called ``energy resolution function''. Actually, its internal representation
in the software is a smearing matrix\index{norm}{Energy resolution}\index{norm}{Smear matrix}. The function $\epsilon_\beta^{\text{IT}}(E')$ will later be refered to as ``post-smearing efficiencies'', since it will allow us to define cuts and threshold functions {\em after} the smearing is performed, \ie, as function of $E'$. The detailed definition and initialization of the energy resolution function is described in \Sec~\ref{sec:energy}.

@@

\index{aedl}{{\tt energy}|(}
The definition and implementation of the energy resolution function is 
rather sophisticated in \GLOBES . In particular, since this aspect is 
 time-critical, several algorithms are implemented. The choice of the 
proper algorithm depends on the experiment and the frequencies of the
 involved neutrino oscillations.

@@

This algorithm is the simplest of the built-in algorithms for the evaluation
of \eq~(\ref{eq:simple_int}). It is applicable to most of the
 accelerator-based experiments which can be simulated with \GLOBES .

>>>>

\index{norm}{AEDL@\AEDL|(}
\part{The Abstract Experiment Definition Language -- AEDL}
\label{part:2}

@@

The goal of AEDL is to describe a large number of complex and very 
different experiments by a limited number of parameters. It allows a
representation of very different setups within one data structure, and thus implements universal rate and $\chi^2$ computation methods. For experiment simulations, usually a new piece of code is written and compiled
for each different experiment. In many cases, even parameter changes, such as
the number of bins, require the recompilation of the source code. 
However, such a technique soon reaches its limits when the simulated experiments are rather complex, or more than one type of experiment is studied simultaneously. Furthermore, it is very difficult to verify the correctness of the obtained results, since every time a new piece of code is added to 
deal with a new experiment type, new errors will be introduced.

Thus, a general and flexible experiment description language is needed.  
The description of a neutrino experiment can be split into three parts: Source, oscillation, and detection. The neutrino sources within \GLOBES\ 
are assumed to be stationary point sources, where each experiment has only 
one source. This restricts the classes of neutrino sources which can be studied with \GLOBES :
\begin{itemize}
\item
 Experiments using many point-like sources can only be approximated. One example are reactor experiments using many distant reactor blocks.
\item
 Geometrical effects of a source distribution, such as in the sun or the atmosphere, can not be described.
\item
 Sources with a physically significant time dependency  can not be studied, such as  supernov\ae. It is, however, possible
to study beams with bunch structure, since the time dependence of the
neutrino source is physically only important to suppress backgrounds. 
\end{itemize}

The description of the neutrino oscillation physics is, at least numerically, relatively simple. We use the {\em evolution operator method}~\cite{Ohlsson:1999um}  to compute the neutrino oscillation probabilities and divide the matter density profile into layers of constant matter density. For each of these layers, the Hamiltonian in matter is diagonalized in order to propagate the neutrino transition amplitudes. Finally, the transition probability is obtained by the absolute square of the total neutrino transition amplitudes. Depending on the precision of the studied experiment, this approach turns out to be precise enough in Earth matter even if only a small number matter density steps is used. Since we allow an uncertainty of the matter density profile, it is, in fact, in most cases sufficient to consider only one density step with the average matter density together with a matter density uncertainty~\cite{Ohlsson:2003ip}. Note that this approach may not be applicable to quickly varying extraterrestrial matter density profiles.

@@

Experiments are in \GLOBES\ defined by the Abstract Experiment Definition Language (AEDL). The experiment definition is written into a text file using the AEDL syntax. Currently, a number of pre-defined experiment definition files are provided with \GLOBES , which have to be modified manually in order to define new experiments.  The application software then uses this text file to initialize the experiment, where other secondary files might read for source fluxes, cross sections \etc . In this section, we show the definition of a very simple neutrino factory in AEDL, where we do not go into details. In the next chapter, we will discuss each of the individual steps in detail.

@@

originating from stored $\mu^+$'s. This achieved by setting the {\tt @builtin} variable to $1$. Next, we specify the muon energy to be $50\,\mathrm{GeV}$ by the {\tt @parent\_energy} variable. We assume 

@@

{\tt \tb @errordim\_sys\_on = 0}\\
{\tt \tb @errordim\_sys\_off = 2}\\

@@

simultaneously. It is only useful if there is more than one background component, otherwise it is usually $1$. The background error variable is defined such as the signal error variable, \ie\ we have a $5\%$ background uncertainty and a very small energy calibration error. The ``error dimension variable'' {\tt @errordim\_sys\_X} selects how the systematical errors are treated (\cf, \Tab~\ref{tab:error_dim}).

@@

We now give a short introduction to the syntax of AEDL.
 The first eight characters have to be {\tt \%!GLoBES}
in order to avoid parsing megabytes of chunk
 and producing thousands of error messages. 
%
Comments can be used such as in C:
\begin{quote}
{\tt /* This starts a comment\\
 and here the comment ends */
}
\end{quote}
There are pre-defined variables which all start with {\tt \$}. Their range
is also checked. For example,  {\tt 
\$bins} can be only between $0$ and $500$.\footnote{The upper limit is 
only there for safety reasons, the memory is allocated dynamically.} If one uses a {\tt float} quantity where  an {\tt int} is expected, the {\tt float} will be converted to an {\tt int} in the same way as in C.  For example, we have scalar variables
\begin{quote}
{\tt
\$bins = 10\\
\$baseline = 1200.0
}
\end{quote}
and simple lists
\begin{quote}
{\tt
\$densitytab=\{1.0,2.2343,3.3432\} 
}
\end{quote}
%
Since there are often groups of data which we want to refer to later,
environments can be used. This is illustrated 
with the channel definition part:
\begin{quote}
{\tt channel(\#ch1)<\\
\tb  $\ldots$\\
>
}
\end{quote}
The first part is the type of environment, which is {\tt channel} here. 
There are the following types of environment in AEDL:
\begin{quote}
{\tt flux\\
cross\\
channel\\
energy\\
rule
}
\end{quote}
Besides the environment type, there is a user-defined name 
beginning with {\tt \#}
in the above example: {\tt \#ch1}. It can be used later to refer to the 
channel defined in {\tt <$\ldots$>}. Those names are so-called 
``automatic variables'' and have to start with {\tt \#}. Note that these names have to be unique and can only be refered to after their definition.
However, similar to C, one can give a declaration without definition before:
\begin{quote}
{\tt    channel(\#ch2)<>}
\end{quote}
Now one can refer to the name {\tt \#ch2}, while the actual channel definition comes later. The internal representation of this automatic
variable is a number, which obtains its value from a counter for each type of environment. For example, for {\tt channel} the counter is {\tt numofchannels}. The counter keeps track of how many different names 
there are for one type of environment, which means that it counts the number of channels, rules, energy resolution functions \etc . Thus, the automatic
variables are numbered in the order of their definition, and the number
can later be used to refer to them in the C code (from $0$ to {\tt numof...}$-1$). In order to facilitate the the mapping from names in AEDL to indices
in C there are two functions \GLB{NameToValue} and \GLB{ValueToName} which
make this transition (see \Sec~\ref{sec:aedl_names}, 
page~\pageref{sec:aedl_names}).

Within each environment type, there are several 
variables beginning with {\tt @}, which can only be used within the 
appropriate type of environment. In many cases, 
they have a special syntax, such as {\tt @channel}.

\index{aedl}{NEXT@{\tt \#NEXT\#}}
If you want to have several experiments in one file, separate the different
 experiments by 
\begin{quote}
{\tt    \#NEXT\#}
\end{quote}
This command resets the counters for channels, rules, fluxes, cross section 
and energy resolution environments. All variables have their scope limited 
by either {\tt \%!GLoBES, \#NEXT\#} or {\tt EOF}.  This allows 
a consistent treatment of various experiments in one file.

\index{aedl}{inlcude@{\tt include}}
As another feature of AEDL one can use include files with the {\tt include} command. Includes can be nested up to {\tt MAX\_INCLUSION\_DEPTH}, which is currently set to $10$. Error reporting works 
 for nested includes, too. The included file is not required to begin 
 with {\tt \%!GLoBES} to facilitate cut and paste:
\begin{quote}
{\tt include "./file\_1"}
\end{quote}
With this include mechanism, one can use constructions such as 
\begin{quote}
{\tt    include  "NuFact.gls"\\
        \#NEXT\#\\
        include   "JHFHK.gls"
}
\end{quote}
in order to initialize a combined analysis of the experiments defined in the files {\tt NuFact.gls} and {\tt JHFHK.gls}. Note that one has 
to use quotation marks for filenames in \AEDL.\index{norm}{File names}
Even if one uses the
automatic variable {\tt \#CC} in both experiments, 
but the cross section data are different (for example, because of different target nuclei), the correct 
cross section data will be applied to each of the experiments. 
Note that, alternatively, one can 
also load both files successively by two separate calls of 
\GLB{InitExperiment}.

Furthermore, one can define constants such as
\begin{quote}
{\tt
Pi = 3.141
}
\end{quote}
These constants can not only be defined within one \AEDL\ file, but also
by the calling C program, which allows to use a simple but powerful variable
substitution mechanism as described in~\Sec~\ref{sec:aedlparams}.
\index{norm}{AEDL@\AEDL!external parameters}

@@

As we have disussed before, \GLOBES\ can only deal with point sources. Thus,  it is not possible to study effects from the finite size of the neutrino production region, such as in the sun or in reactor experiments with many
neutrino sources (\eg, KamLAND). Therefore, a neutrino source in \GLOBES\ can, in general, be characterized by the flux spectrum for each neutrino flavor, the CP sign (neutrinos or antineutrinos), and the total luminosity
of the source.

Before we come to the definition of the source properties, let us discuss
the total integrated luminosity of the experiment. In \GLOBES , the total number of events is in general proportional to the product of
\begin{equation}
\mathrm{Fid.~detector~mass}\,\left[\mathrm{kt/t}\right]\times 
\mathrm{Running~time} \,\left[\mathrm{yr}\right]\times\left\{ \begin{array}{c}
\mathrm{Source~power}\,\left[\mathrm{MW/GW}\right]\\
\mathrm{Useful~muon~decays}\,\left[\mathrm{yr}^{-1}\right]
\end{array}\right.\,.
\end{equation}
Thus, the source power corresponds to either the amount of energy produced per time frame in the target (such as for nuclear reactors or sources based on pion decay), or the useful muon decays per time frame (neutrino factories). In addition, the definition of the source power makes only sense together with the flux normalization, the running time, and fiducial detector mass in order to define the total integrated luminosity. Therefore, one can, in principle, use arbitrary units for these components as long as their product gives the wanted neutrino flux. However, it is
recommended to use normalizations such that the source power units are $\mathrm{MW}$ for a proton-based beam, and $\mathrm{GW}_\mathrm{thermal}$ for a reactor experiment. Correspondingly, the detector mass units should be kilotons for a proton-based beam, and tons for a reactor experiment. In any case it is a good
idea to document the choices made by the user by corresponding comments
in \AEDL.

@@

\index{aedl}{{\tt flux}!flux file@{\tt "@flux\_file}}\index{norm}{Flux!file}
The software assumes that the given flux file has seven columns and
501 lines with equidistant energies. The format is:
\begin{quotation}
$ E\quad
\Phi_{\nu_e}\quad
\Phi_{\nu_\mu}\quad
\Phi_{\nu_\tau}\quad
\Phi_{\bar\nu_e}\quad
\Phi_{\bar\nu_\mu}\quad
\Phi_{\bar\nu_\tau}$
\end{quotation}
In order to access fluxes at arbitrary energies, linear interpolation 
is used by the software. In general, it is advisable to provide the flux between {\tt \$sampling\_min} and {\tt \$sampling\_max} (\cf, \Sec~\ref{sec:energy}), since these values are used by the software. However, if the energy leaves the range of values given in the file, zero is returned.
The the columns for the fluxes for unused flavours have to be filled all the 
same,\eg\ with zeros.

The flux files accept one-line comments, which start
with {\tt \#} and end with the linefeed character `\verb+\n+', they are
not counted as a line and their content is discarded. This comments
are useful to provide meta information about the fluxes like units
or the origin of the information. This is also the default method
to point the user to the references he/she should to cite when using
a particular flux.\index{norm}{Flux!file!comments in}
\index{norm}{Referencing!flux data}

@@

Furthermore, the matter density profile along the baseline
has to be specified. The simplest matter density profile is a constant matter density profile with the average matter density from the PREM~\cite{Stacey} onion shell model of the earth\index{norm}{PREM| \see{Matter density}}
\index{norm}{Matter density!of the earth}:
\begin{quote}
{\tt \$profiletype=1 }
\end{quote}
%
If your using this option please cite reference~\cite{Stacey}\index{norm}{Referencing!matter profile data}.

For a better approximation of the realistic earth matter density profile, one can use an arbitrary number of equidistant steps of the PREM profile:
\begin{quote}
{\tt \$profiletype=2 } \\
{\tt \$densitysteps=20 }
\end{quote}
Note that the value of {\tt \$densitysteps} is time-critical, since the
computation time of oscillation probabilities is directly 
proportional to the number of layers.
%
As a  third possibility, one can specify the matter density profile 
manually with a list of thicknesses and densities of the matter density layers. This example uses two density steps with two different densities:
\begin{quote}
{\tt \$profiletype=3 } \\
{\tt \$densitytab=\{2.8, 3.5\}}\\
{\tt \$lengthtab=\{1000.0, 2000.0\}}
\end{quote}
It is important that both lists have the same length and that the  thicknesses given in  {\tt \$lengthtab} add up to the length of
the baseline, which does not have to be explicitely specified anymore. In addition, matter densities are always given in $g/cm^3$.
%
This approach can also be used for a constant matter density profile with
a specific matter density:
\begin{quote}
{\tt \$profiletype=3 } \\
{\tt \$densitytab=\{3.5\}}\\
{\tt \$lengthtab=\{3000.0\}}
\end{quote}
The possible options for matter density profiles are summarized in \Tab~\ref{tab:profiletypes}.

@@

Cross sections\index{norm}{Cross section} will later be used as part of the 
channel definition (see \Sec~\ref{sec:channel}). Similar to the source 
fluxes, they are provided by the user as a data file:
\index{aedl}{{\tt cross}}
\index{aedl}{{\tt cross}!cross file@{\tt  "@cross\_file}}
\begin{quote}
{\tt cross(\#name)<}\\
{\tt \tb @cross\_file ="user\_file\_1.dat"}\\
{\tt >}
\end{quote}  
This cross section can later be refered to by {\tt \#name}.

Cross sections are in \GLOBES\ given as differential cross section per energy:
\begin{equation}
\hat\sigma(E)=\sigma(E)/E\,\left[ 10^{-38}\,
\frac{\mathrm{cm}^2}{\mathrm{GeV}^2} \right]
\end{equation}
The software assumes that the cross section files are text files with 
seven columns and $1001$ lines of the form
\index{norm}{Cross section!file}
\index{aedl}{{\tt cross}!cross file@{\tt  "@cross\_file}}
\begin{quotation}
$\mathrm{log}_{10} E\quad
\hat\sigma_{\nu_e}\quad
\hat\sigma_{\nu_\mu}\quad
\hat\sigma_{\nu_\tau}\quad
\hat\sigma_{\bar\nu_e}\quad
\hat\sigma_{\bar\nu_\mu}\quad
\hat\sigma_{\bar\nu_\tau}$
\end{quotation}
Here the logarithms of the energy values have to be equidistant. For 
arbitrary energies, linear interpolation is used. If the energy leaves the
range of values given in the file, $0.0$ will be assumed. In general, it is advisable to provide the cross sections in the range between {\tt \$sampling\_min} and {\tt \$sampling\_max} (\cf, \Sec~\ref{sec:energy}).
Unused cross sections have to be filled with zeros, and can not
 be just omitted.

Like the flux files, the cross section files 
accept one-line comments, which start
with {\tt \#} and end with the linefeed character `\verb+\n+', they are
not counted as a line and their content is discarded. This comments
are useful to provide meta information about the cross sections like units
or the origin of the information. This is also the default method
to point the user to the references he/she should to cite when using
a particular cross section.\index{norm}{Cross section!file!comments in}
\index{norm}{Referencing!cross section data}

@@

\index{aedl}{{\tt channel}|(}
Channels\index{norm}{Channel} in \GLOBES\ represent an intermediate level 
between the pure oscillation physics given by the oscillation probability
$P_{\alpha\beta}$, and the total event rates composed of signal and 
background. A channel describes the path from one
initial neutrino flavor in the source to the event rates in the detector for one specific interaction type (IT) and final flavor.  
Therefore, a channel contains the description of the 
initial neutrino flavor, its CP eigenvalue 
(neutrino or antineutrino)\footnote{Currently \GLOBES\ does not support lepton
number violating transitions, \ie\ no transitions from neutrino to 
antineutrino (or vice versa) are considered.}, 
the detected neutrino flavor, the interaction cross sections for the chosen interaction type, and the energy resolution function of the detector.

@@

Since it is a lot of effort to solve this double integral numerically,
we split up the two integrations. First, we evaluate the integral over
$\hat{E}$, where the only terms containing $\hat{E}$ are
$k_\beta^{\text{IT}}(E-\hat{E})$,  $ T_\beta(\hat{E})$, and 
$ V_\beta(\hat{E}-E')$. We define:
\begin{eqnarray}
\label{eq:e_res} 
R_\beta^{\text{IT}}(E,E')\,\epsilon_\beta^{\text{IT}}(E')
 \equiv
\int\limits_0^\infty d\hat{E} \quad T_\beta(\hat{E})\,k_\beta^{\text{IT}}(E-\hat{E})
\,V_\beta(\hat{E}-E')\,. 
\end{eqnarray}
Thus, $R_\beta^{\text{IT}}(E,E')$ describes the energy response of 
the detector, \ie , a neutrino with a (true) energy $E$ is reconstructed
with an energy between $E'$ and $E'+dE'$ with a probability
$R_\beta^{\text{IT}}(E,E') dE'$. The function $R(E,E')$ is also often called ``energy resolution function''. Actually, its internal representation
in the software is a smearing matrix\index{norm}{Energy!resolution}\index{norm}{Smear matrix}. The function $\epsilon_\beta^{\text{IT}}(E')$ will later be refered to as ``post-smearing efficiencies'', since it will allow us to define cuts and threshold functions {\em after} the smearing is performed, \ie, as function of $E'$. The detailed definition and initialization of the energy resolution function is described in \Sec~\ref{sec:energy}.

@@

\index{aedl}{{\tt energy}|(}
\index{norm}{Energy!resolution|(}
The definition and implementation of the energy resolution function is 
rather sophisticated in \GLOBES . In particular, the choice of the 
proper parameters depends on the experiment and the frequencies of the
 involved neutrino oscillations. This choice also greatly influences the
speed of the calculation.

@@

This algorithm is the simplest of the built-in algorithms for the evaluation
of \eq~(\ref{eq:simple_int}). It is applicable to most of the
experiments which can be simulated with \GLOBES .
